---
title: "Geospatial Data Science"
Authors:
- Bruno Conte
format: html
toc: true
---

```{r}
library(tidyverse)
library(readxl)
library(here)
library(leaflet)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(stringi)
```

```{r}
# Read municipios_completo sheet from Municipios.xlsx
FILE_MUNICIPIOS <- here("Municipios.xlsx")

# Read the "municipios_completo" sheet using tidyverse and rename columns
df_municipios_completo <- read_excel(
  FILE_MUNICIPIOS, 
  sheet = "municipios_completo") %>%
  rename(
    Municipio = nomMun, 
    Departamento = nomDepto) %>%
  mutate(
    Departamento = stri_trans_general(str_trim(Departamento), "Latin-ASCII"),
    Municipio = stri_trans_general(str_trim(Municipio), "Latin-ASCII")
  )

# View the dataframe
df_municipios_completo
```

```{r}
# # Create interactive map with dots for each municipality
# leaflet(df_municipios_completo) %>%
#   addTiles() %>%
#   addCircleMarkers(
#     ~lon, ~lat,
#     radius = 3,
#     color = "blue",
#     fillOpacity = 0.6,
#     popup = ~paste0("Municipio: ", Municipio, "<br>Lat: ", lat, "<br>Lon: ", lon)
#   )
```

```{r}
# Read Municipio sheet from Remesas.xlsx
FILE_REMESAS <- here("Remesas.xlsx")

# Read the "Municipio" sheet using tidyverse
df_remesas_municipio <- read_excel(
  FILE_REMESAS, 
  sheet = "Municipio", 
  skip = 0) %>%
  mutate(
    Departamento = stri_trans_general(str_trim(Departamento), "Latin-ASCII"),
    Municipio = stri_trans_general(str_trim(Municipio), "Latin-ASCII")
  )
df_remesas_municipio
```

```{r}
# Perform the join
df_remesas_municipio <- df_remesas_municipio %>%
  left_join(
    df_municipios_completo %>% select(Departamento, Municipio, lat, lon),
    by = c("Departamento" = "Departamento", "Municipio" = "Municipio")
  )

# Display the result with new columns
df_remesas_municipio
```

```{r}
# Report NA counts for latitude and longitude
na_counts <- df_remesas_municipio %>%
  summarise(
    lat_na = sum(is.na(lat)),
    lon_na = sum(is.na(lon)),
    total_rows = n()
  )

print(na_counts)

cat("\nFirst rows with missing lat or lon:\n")
print(
  df_remesas_municipio %>%
    filter(is.na(lat) | is.na(lon)) %>%
  select(Departamento, Municipio, lat, lon)
)
```

# Remittances in Honduras: Geospatial Analysis by Department

[Source: Central Bank of Honduras (BCH)](https://www.bch.hn/estadisticas/remesas-familiares)

```{r}
# Get Honduras boundaries
honduras <- ne_states(
  country = "Honduras", 
  returnclass = "sf")
honduras_country <- ne_countries(
  scale = "medium", 
  country = "Honduras", 
  returnclass = "sf")
```

```{r}
df_dept <- df_remesas_municipio %>%
  group_by(Departamento) %>%
  summarise(monto_sum = sum(.data[["Monto"]], na.rm = TRUE), .groups = "drop")

# compute national total and proportion
total_remesas <- sum(df_dept$monto_sum, na.rm = TRUE)
df_dept <- df_dept %>% 
  mutate(prop = monto_sum / total_remesas)
df_dept

honduras_dept_sf <- honduras %>%
  mutate(
    Departamento_shape = stri_trans_general(str_trim(.data[["name"]]), "Latin-ASCII")
  )

# normalize df_dept names for join
df_dept <- df_dept %>% 
  mutate(
    Departamento = stri_trans_general(
      str_trim(Departamento), "Latin-ASCII"))

# join shapes with aggregated data
honduras_dept_sf <- honduras_dept_sf %>%
  left_join(df_dept, by = c("Departamento_shape" = "Departamento"))

library(viridis)

ggplot(honduras_dept_sf) +
  geom_sf(aes(
    fill = prop), color = "white", linewidth = 0.2) +
  scale_fill_gradient(low = "#deebf7", high = "#08306b", labels = scales::percent_format(accuracy = 0.1), name = "Share of\nTotal Remesas") +
  theme_minimal() +
  labs(title = "Proportion of Remittances by Department",
       subtitle = "Share of national remittances received by department",
       caption = paste0("Total remittances: ", scales::comma(total_remesas))) +
  theme(legend.position = "right")
```

# Remmitances in Honduras: Geospatial Analysis by Municipality

```{r}
# Create static map with ggplot including boundaries and point size = summed Monto
ggplot() +
  geom_sf(
    data = honduras_country, 
    fill = "white", 
    color = "black", 
    linewidth = 1) +
  geom_sf(
    data = honduras, 
    fill = NA, 
    color = "gray50", 
    linewidth = 0.5) +
  geom_point(
    data = df_remesas_municipio %>%
      group_by(Departamento, Municipio, lat, lon) %>%
      summarise(Monto = sum(Monto, na.rm = TRUE)),
    aes(x = lon, y = lat, size = Monto/1000000),
    color = "blue", 
    alpha = 0.6) +
  scale_size_area(
    max_size = 10, 
    name = "USD Amount\nMillions") +
  theme_minimal() +
  labs(
    title = "Municipalities: Size by Sum of USD Amount",
    caption = "Note: The data was obtained by a query to the Central Bank of Honduras (transparencia@bch.hn)",
    x = NULL,
    y = NULL
  ) +
  coord_sf() +
  theme(plot.caption = element_text(hjust = 0))
```

The reception of remittances in Honduras is highly concentrated in the departments of Francisco Morazán and Cortés, which includes the major cities of Tegucigalpa and San Pedro Sula. This is evident from the larger point size representing the sum of remittances in that area.

The second map highlights the spatial distribution of remittance reception across different municipalities, indicating that certain areas are more reliant on remittances than others.


# Business Directory Summary (DEE 2024)

[Source: DEE 2024 Excel file](https://ine.gob.hn/dee/)

```{python}
import polars as pl

# 1. Read Excel
df_raw = pl.read_excel("data/BASE-DEE-2024.xlsx")

# First data row contains actual column headers
new_headers = [str(v) for v in df_raw.row(0)]
df = df_raw.slice(1).rename(dict(zip(df_raw.columns, new_headers)))

# 2. Group-by summary (count)
group_cols = [
    "COD. DEPARTAMENTO",
    "COD. MUNICIPIO",
    "CÓDICO ",
    "TIPO DE EMPRESA",
    "CÓDIGO GENERAL",
]

summary = (
    df
    .group_by(group_cols)
    .agg(pl.len().alias("N"))
    .sort(group_cols)
    .drop_nulls()
    .rename({
        "COD. DEPARTAMENTO": "codDepto",
        "COD. MUNICIPIO": "codMun",
        "CÓDICO ": "codDeptoMun",
        "TIPO DE EMPRESA": "Size",
        "CÓDIGO GENERAL": "Activity",
    })
)

print(f"Summary shape: {summary.shape}")
print(summary)

# ── 3. Export to CSV ───────────────────────────────────────────
summary.write_csv("data/summary_dee_2024.csv")
print("CSV saved to: data/summary_dee_2024.csv")
```

```{r}
# Read DEE summary and aggregate total businesses per municipality
df_dee <- read_csv("data/summary_dee_2024.csv") %>%
  group_by(codDeptoMun) %>%
  summarise(N = sum(N, na.rm = TRUE), .groups = "drop")

# Join with municipality coordinates
df_dee <- df_dee %>%
  left_join(
    df_municipios_completo %>% select(codDeptoMun, Departamento, Municipio, lat, lon),
    by = "codDeptoMun"
  )

# Map: Business establishments per municipality
ggplot() +
  geom_sf(
    data = honduras_country,
    fill = "white",
    color = "black",
    linewidth = 1) +
  geom_sf(
    data = honduras,
    fill = NA,
    color = "gray50",
    linewidth = 0.5) +
  geom_point(
    data = df_dee,
    aes(x = lon, y = lat, size = N),
    color = "darkgreen",
    alpha = 0.6) +
  scale_size_area(
    max_size = 10,
    name = "Number of\nEstablishments") +
  theme_minimal() +
  labs(
    title = "Business Establishments by Municipality (DEE 2024)",
    caption = "Source: INE Honduras - Directorio de Establecimientos Económicos 2024",
    x = NULL,
    y = NULL
  ) +
  coord_sf() +
  theme(plot.caption = element_text(hjust = 0))
```

# Population Density in Honduras: Geospatial Analysis by Department and Municipality

[Source: INE Honduras / UNFPA — Humanitarian Data Exchange (HDX)](https://data.humdata.org/dataset/6203b8df-ef66-4e55-baba-6d8547438a77)

```{r}
# Read HDX population data (municipality level, 2024 projections)
df_pop <- read_csv("data/hnd_admpop_adm2_2024.csv") %>%
  mutate(
    # Extract the 4-digit code from ADM2_PCODE (e.g., HN0102 -> 0102)
    codDeptoMun = str_sub(ADM2_PCODE, 3, 6)
  ) %>%
  select(ADM1_ES, ADM2_ES, ADM2_PCODE, codDeptoMun, T_TL)

# Join with municipality coordinates
df_pop <- df_pop %>%
  left_join(
    df_municipios_completo %>% select(codDeptoMun, lat, lon),
    by = "codDeptoMun"
  )

# Map: Population by municipality
ggplot() +
  geom_sf(
    data = honduras_country,
    fill = "white",
    color = "black",
    linewidth = 1) +
  geom_sf(
    data = honduras,
    fill = NA,
    color = "gray50",
    linewidth = 0.5) +
  geom_point(
    data = df_pop,
    aes(x = lon, y = lat, size = T_TL / 1000),
    color = "darkred",
    alpha = 0.6) +
  scale_size_area(
    max_size = 12,
    name = "Population\n(thousands)") +
  theme_minimal() +
  labs(
    title = "Population by Municipality (2024 Projections)",
    caption = "Source: INE Honduras / UNFPA via Humanitarian Data Exchange (HDX)",
    x = NULL,
    y = NULL
  ) +
  coord_sf() +
  theme(plot.caption = element_text(hjust = 0))
```
